\section{Selected Inference Algorithms for TKF92}
\label{sec:tkf-inference}

This section describes four inference algorithms that operate on
TKF91 / TKF92 models and on phylogenetic trees:
the \emph{Fast Statistical Alignment} (FSA) framework with a
Newton-step time-optimization on the per-cherry expected
log-likelihood (Section~\ref{sec:fsa});
the \emph{Beam Search Ancestral Reconstruction} (BeamASR) algorithm
(Section~\ref{sec:beam-ancestor});
the \emph{Variational Ancestral Reconstruction} (VarAnc) algorithm
that operates on a fixed MSA with known gap structure
(Section~\ref{sec:varanc});
and its stochastic-variational training extension
\emph{svi-VarAnc} (Section~\ref{sec:svi-varanc}).

\subsection{Fast Statistical Alignment (FSA)}
\label{sec:fsa}

Given a set of sequences and a phylogenetic tree with branch-specific
TKF91 or TKF92 pair HMMs, we construct a multiple sequence alignment
using the \emph{sequence annealing} approach of \cite{BradleyEtAl2009},
to which we refer the reader for a full description of the algorithm.

Briefly, the method proceeds as follows.
For each pair of sequences $(x,y)$ in a selected subset
(either all $\binom{N}{2}$ pairs or an $O(N \log N)$ Erd\H{o}s--R\'enyi sample),
we compute pairwise residue alignment posteriors $P(x_i \sim y_j)$
by running the Forward-Backward algorithm on the pair HMM at an
optimized evolutionary time $\hat{\tau}$.

\paragraph{Per-pair time optimization (NR step).}
The time $\hat{\tau}$ is found by Newton--Raphson optimization
of the expected log-likelihood (the ``NR step''):
\begin{align}
\hat{\tau} &= \operatorname*{argmax}_\tau \;
   \mathbb{E}_{P(\pi | x,y,\tau_0)} \bigl[ \log P(x,y,\pi \mid \tau) \bigr]
\label{eq:fsa-nr}
\end{align}
where $\pi$ ranges over alignment paths and $\tau_0$ is an initial
estimate.  This expectation is computed from Forward--Backward expected
counts at $\tau_0$, and---because the TKF92 expected log-likelihood is
quadratic-like in $\tau$ in a neighbourhood of the maximum, with
analytic derivatives via the BDI score identities of
Section~\ref{sec:bdi}---the Newton iteration typically converges in
3--5 steps.
This time-maximisation differs from the original FSA approach of
\cite{BradleyEtAl2009}, which attempts to optimize all model parameters
via unregularized EM for every pair, and consequently must terminate
the EM recursion early to avoid instability.  Restricting the optimization
to the per-pair time $\tau$, while keeping the model parameters fixed
at the population estimate, is both cheaper and more stable.

\paragraph{Sequence annealing.}
The pairwise posteriors are then assembled into a multiple alignment
by the greedy sequence annealing procedure of \cite{BradleyEtAl2009},
which iteratively merges alignment columns to maximize a sum-of-pairs
posterior objective.

\subsection{Beam Search Ancestral Sequence Reconstruction (BeamASR)}
\label{sec:beam-ancestor}

We now describe an alternative progressive reconstruction method
that finds the maximum-likelihood ancestral sequence at each internal node
by beam search, without materializing the full composite automaton.

At each internal node $v$ with children $l,r$ and observed descendant sequences
$c_l, c_r$, we seek
\begin{align}
\hat{a}_v &= \operatorname*{argmax}_{a} \bigl[ \log P(a, c_l \mid B_l) + \log P(a, c_r \mid B_r) - \log P(a \mid R) \bigr]
\label{eq:beam-ancestor}
\end{align}
where $P(a, c_k \mid B_k)$ is the pair HMM forward probability on branch $k$
and $P(a \mid R)$ is the singlet probability under the root generator,
subtracted to avoid double-counting the prior on~$a$.

\subsubsection{Incremental forward profiles}

Since the branches are conditionally independent given the ancestor,
we can evaluate~(\ref{eq:beam-ancestor}) by maintaining
\emph{incremental forward profiles}: for each branch $k$, a 1D forward table
$F_k[i, q]$ giving the log-probability that descendant positions $1,\ldots,i$
have been emitted and the branch machine is in state $q$,
given ancestor positions $1,\ldots,j$ processed so far.

Each ancestor character extends both profiles independently in $O(L_k)$ time
per branch, where $L_k = |c_k|$.

\subsubsection{Beam search}

The ancestor sequence $\hat{a}_v$ is built left-to-right by beam search.
At each position $j$, the beam maintains $B$ candidate partial ancestors.
For each candidate and each alphabet character $\sigma$:
\begin{enumerate}
\item Extend both branch profiles by one ancestor character $\sigma$,
  comprising a \emph{match/delete phase} (the ancestor emits $\sigma$,
  descendant positions advance via M or D transitions)
  and an \emph{insertion phase} (descendant-only insertions following the
  ancestor emission).
\item Update the singlet forward score for $\sigma$.
\item Score the extension:
  $\Delta(j, \sigma) = \Delta F_{l} + \Delta F_{r} - \Delta_{\mathrm{singlet}}$.
\end{enumerate}
The top $B$ extensions (by cumulative score) are retained.
Total cost per node is $O(K \cdot B \cdot A \cdot (L_l + L_r))$
where $K = |\hat{a}_v|$ and $A$ is the alphabet size.

\subsubsection{Insertion phase via associative scan}

The insertion recurrence within each branch profile has the form
\begin{align}
x_{i+1} &= \operatorname{logsumexp}\bigl(A_{II} \, x_i,\; b_i\bigr) + e_i
\label{eq:insert-recurrence}
\end{align}
where $A_{II}$ is the I-to-I log-transition submatrix, $b_i$ collects
transitions into insertion states from M and D, and $e_i$ is the emission score.
This is a log-semiring affine recurrence, parallelizable via
an associative scan with operator
\[
(A_1, b_1) \oplus (A_2, b_2) = (A_2 \otimes A_1,\; \operatorname{logsumexp}(A_2 \, b_1,\; b_2))
\]
where $\otimes$ denotes log-semiring matrix multiplication.
This reduces the insertion phase from $O(L)$ sequential depth to $O(\log L)$.

\subsubsection{Supported model types}

The beam search interface is generic over the pair HMM used on each branch.
For the present (TKF) paper we specialize to:
\begin{enumerate}
\item \textbf{TKF91} --- order-0, 5-state pair HMM.
\item \textbf{TKF92} --- order-0, 5-state pair HMM with fragment self-loops
  (the standard model of Section~\ref{sec:tkf92-machines}).
\end{enumerate}
The same beam-search framework extends without algorithmic change to
hierarchical TKF extensions in which each branch carries a
larger latent-state pair HMM \cite{LargeHolmes2026}; we
do not pursue that direction here.


\subsection{Variational Ancestral Reconstruction (VarAnc)}
\label{sec:varanc}

Given an MSA with known gap structure---i.e., for each internal node
$v$ and MSA column $c$, we know whether $v$ is present (ungapped) or
absent (gapped)---but unknown ancestral residue identities, we
optimize a variational lower bound (ELBO) on the marginal likelihood.
Our approach is a ``product of trees'' structured variational
approximation~\cite{JojicEtAl2004}.

\subsubsection{Model structure}

The joint distribution over observed leaf sequences $\mathbf{y}$
and ancestral sequences $\mathbf{a}$ factorises over tree edges.
With the alignment structure fixed, the log-likelihood decomposes into
contributions from consecutive alignment-column pairs along each edge $e = (u,v)$:
\begin{align}
\log P(\mathbf{y}, \mathbf{a})
 &= \sum_{e \in \mathrm{edges}} \sum_{k=1}^{K_e}
    \log w_e\!\bigl(\mathrm{type}_{k-1},\, \mathrm{type}_k,\, \mathrm{ctx}_{k-1},\, \mathrm{ctx}_k\bigr)
\label{eq:lik-edges-tkf}
\end{align}
where $w_e$ is the TKF92 pair-HMM transition weight on edge $e$
(depending on the previous and current column types and context characters,
plus the branch length).
The first and last columns contribute start ($\sta$) and end ($\fin$) transition factors.

The key computational challenge is that even the order-0 TKF92 transitions
couple adjacent alignment columns through the fragment-extension
self-loops: the pair-HMM weight at column $c$ depends on the
ancestral and descendant characters at both column $c$ and its predecessor.
This makes the full graphical model a tree (phylogeny) $\times$ chain (sequence)
grid with undirected cycles, and exact inference is intractable.

\subsubsection{Product-of-trees variational distribution}

Following~\cite{JojicEtAl2004}, we factorize the variational posterior
over MSA columns while retaining the full tree joint at each column:
\begin{align}
q(\mathbf{a}) &= \prod_{c=1}^{L} q_c\!\bigl(\mathbf{h}_c\bigr)
\label{eq:var-q-pot-tkf}
\end{align}
where $\mathbf{h}_c = \{a_{v,c} : v \in \mathrm{ungapped}(c)\}$ denotes the set of
all ancestral characters at MSA column $c$,
and each $q_c$ is a tree-structured distribution over the characters at the
internal nodes that are ungapped at column $c$.
Leaf characters are observed (clamped).

This approximation decouples columns in $q$ but preserves the
full parent-child correlation structure within each column.
In particular, the pairwise joint marginal
$q_c^{(u,v)}(a_u, a_v)$ on each tree edge is available exactly from
the Felsenstein peeling/unpeeling pass used to represent $q_c$
(Section~\ref{sec:felsenstein-update-tkf}).

\subsubsection{Free energy and ELBO}

The evidence lower bound is
\begin{align}
\mathcal{L}(q) &= \mathbb{E}_q\bigl[\log P(\mathbf{y}, \mathbf{a})\bigr] + H(q)
  \;\geq\; \log P(\mathbf{y})
\label{eq:elbo-tkf}
\end{align}
where $H(q) = -\sum_c \sum_{\mathbf{h}_c} q_c(\mathbf{h}_c) \log q_c(\mathbf{h}_c)$
is the entropy of the product-of-trees distribution.
Since each $q_c$ is tree-structured, its entropy decomposes as
\begin{align}
H(q_c) &= \sum_{v \in \mathrm{ungapped}(c)} H\!\bigl(q_c^{(v)}\bigr)
  - \sum_{e = (u,v)} \mathrm{MI}_c(u, v)
\label{eq:tree-entropy-tkf}
\end{align}
where $H(q_c^{(v)})$ is the marginal entropy at node $v$ and
$\mathrm{MI}_c(u,v)$ is the mutual information between parent $u$ and child $v$
at column $c$, both computable from the Felsenstein marginals.

\subsubsection{Potentials from neighboring columns}

The inter-column coupling enters through the TKF92 pair-HMM transitions.
For each edge $e = (u,v)$ and MSA column $c$ where $e$ has an event of type $\tau_c$,
let $c^-$ denote the predecessor column (the previous column where $e$ had an event)
and $c^+$ the successor column.
The \emph{potential} at column $c$ for edge $e$ receives two contributions:

\paragraph{As-child term (from $c^-$).}
The transition from column $c^-$ to $c$ on edge $e$ depends on the
characters at both columns. Using the pairwise marginal from $q_{c^-}$:
\begin{align}
\log \phi_e^{\mathrm{child}}(a_{u,c}, a_{v,c})
 &= \sum_{a', b'} q_{c^-}^{(u,v)}(a', b')\;
    \log w_e(\tau_{c^-\!}, \tau_c, a', b', a_{u,c}, a_{v,c})
\label{eq:pot-child-tkf}
\end{align}
where the sum over $(a', b')$ uses the \emph{joint} pairwise marginal
$q_{c^-}^{(u,v)}(a', b')$---not the product of independent marginals.
This is the key advantage over mean-field:
the within-tree parent-child correlation at the predecessor column is
preserved exactly.

\paragraph{As-parent term (from $c^+$).}
Symmetrically, column $c$ acts as the predecessor for column $c^+$:
\begin{align}
\log \phi_e^{\mathrm{parent}}(a_{u,c}, a_{v,c})
 &= \sum_{a'', b''} q_{c^+}^{(u,v)}(a'', b'')\;
    \log w_e(\tau_c, \tau_{c^+\!}, a_{u,c}, a_{v,c}, a'', b'')
\label{eq:pot-parent-tkf}
\end{align}

For insert transitions (only the descendant is present at $c$),
the potential reduces to a per-node function $\psi_v(a_{v,c})$.
For delete transitions (only the ancestor is present), it becomes $\psi_u(a_{u,c})$.
For match transitions, it contributes a per-edge potential $\phi_e(a_{u,c}, a_{v,c})$.
The start and end transitions contribute analogous per-node or per-edge terms.

\subsubsection{Felsenstein coordinate ascent}
\label{sec:felsenstein-update-tkf}

Each coordinate ascent step updates $q_c$ for a single MSA column $c$,
holding all other columns fixed.
We accumulate, for each edge $e$ and node $v$ in the tree at column $c$:
\begin{itemize}
\item Per-edge log-potentials
  $\log \phi_e(a_u, a_v) = \log \phi_e^{\mathrm{child}} + \log \phi_e^{\mathrm{parent}}$
  (for match transitions where both endpoints are present).
\item Per-node log-potentials
  $\log \psi_v(a_v)$ (from insert/delete transitions on incident edges,
  plus the root prior $\log \pi(a)$ at the root node).
\end{itemize}

The optimal $q_c$, given the potentials, is the Gibbs distribution on the
tree at column $c$:
\begin{align}
q_c(\mathbf{h}_c) &\propto \prod_v \psi_v(a_v) \prod_{e=(u,v)} \phi_e(a_u, a_v)
  \prod_{\ell \in \mathrm{leaves}} \delta(a_\ell = y_\ell)
\label{eq:q-gibbs-tkf}
\end{align}

Since this is a tree-structured MRF, the normalising constant and all
node and edge marginals can be computed \emph{exactly} by Felsenstein
peeling (postorder) and unpeeling (preorder) in $O(|E| \cdot |\alphabet|^2)$ time.

\paragraph{Peeling (postorder).}
For each node $v$ in postorder, compute the conditional likelihood:
\begin{align}
\mathrm{CL}_v(a) &= \psi_v(a) \prod_{\mathrm{children}\; c}
  \Bigl[\sum_{a_c} \phi_{(v,c)}(a, a_c)\; \mathrm{CL}_c(a_c)\Bigr]
\label{eq:peel-tkf}
\end{align}
with $\mathrm{CL}_\ell(a) = \delta(a = y_\ell)$ for observed leaves.
The log-partition function is $\log Z_c = \log \sum_a \pi(a)\, \mathrm{CL}_{\mathrm{root}}(a)$.

\paragraph{Unpeeling (preorder).}
Propagate top-down to obtain the posterior marginal at each node:
\begin{align}
q_c^{(v)}(a) &\propto \mathrm{CL}_v(a) \cdot \mathrm{msg}_{\mathrm{parent} \to v}(a)
\label{eq:unpeel-marginal-tkf}
\end{align}
and the pairwise marginal on each edge:
\begin{align}
q_c^{(u,v)}(a_u, a_v)
 &\propto \mathrm{msg}_{\mathrm{above}\,u}(a_u) \cdot \phi_{(u,v)}(a_u, a_v) \cdot \mathrm{CL}_v(a_v)
\label{eq:unpeel-pair-tkf}
\end{align}
where the ``message from above $u$'' combines the top-down message to $u$
with $u$'s conditional likelihood excluding child $v$.

\paragraph{Sweep.}
One iteration sweeps through all MSA columns $c = 1, \ldots, L$:
for each column, recompute the potentials from the current neighbor marginals,
run peeling/unpeeling, and store the updated node and edge marginals.

\subsubsection{Properties}

The product-of-trees approximation enjoys the same monotonic convergence
guarantee as mean-field coordinate ascent (each column update minimises
the free energy in its coordinate), with the additional guarantee that
the ELBO is at least as tight as the fully-factored mean-field bound.
This follows because the product-of-trees family \emph{contains} the
mean-field family as a special case (where each $q_c$ is itself fully factored).

\paragraph{Computational cost.}
The dominant cost per sweep is the potential computation:
$O(|E| \cdot L \cdot |\alphabet|^4)$ for Match$\to$Match transitions
(contracting the $(|\alphabet|, |\alphabet|)$ pairwise marginal against the
$(|\alphabet|, |\alphabet|, |\alphabet|, |\alphabet|)$ pair-HMM tensor).
The Felsenstein passes add $O(|E| \cdot L \cdot |\alphabet|^2)$, which is subdominant.


\subsection{Stochastic Variational VarAnc (svi-VarAnc)}
\label{sec:svi-varanc}

The VarAnc algorithm of Section~\ref{sec:varanc} computes a variational
posterior over ancestral residues for a \emph{fixed} TKF92 model
$\theta$ on a \emph{single} fixed-MSA family.  We now extend it to a
training algorithm that updates $\theta$ on a corpus of tree-structured
MSAs.  The construction parallels SVI-BW
(Section~\ref{sec:svi-bw}) but uses an MSA-conditioned VarAnc E-step
in place of the per-pair Pair-HMM Forward--Backward.

\paragraph{Outer EM.}
The outer iteration is a stochastic-EM loop over MSAs.  At each
iteration:
\begin{enumerate}[nosep]
\item Sample a minibatch of MSAs (with their phylogenetic trees,
  obtained e.g.\ by FastTree).
\item For each MSA, run a few sweeps of the VarAnc coordinate-ascent
  algorithm of Section~\ref{sec:felsenstein-update-tkf} to refine
  $q$.  This is the \emph{family-level E-step}.
\item Read out, from the converged $q$ and the pairwise edge
  marginals $q_c^{(u,v)}$, the expected sufficient statistics
  $(B,D,L,M,S,T,F,E,W,U,V)$ for the TKF92 model:
  \begin{itemize}[nosep]
  \item Indel/fragment statistics $(B,D,L,M,S,T,F,E)$ are read
    column-by-column from the per-edge expected event types
    (match/insert/delete) implied by the gap structure, weighted
    by the pairwise edge marginals.  These are the same TKF92
    sufficient statistics defined in Section~\ref{sec:bw-tkf92}.
  \item Substitution statistics $(W,U,V)$ are accumulated by
    contracting the pairwise edge marginal $q_c^{(u,v)}(a_u,a_v)$
    against the bridge expectations
    $\emcount^W_i(a_u,a_v,t_e)$ and $\emcount^U_{ij}(a_u,a_v,t_e)$ at
    each match column.
  \end{itemize}
\item Apply the SVI exponential moving average update
  (equation~(\ref{eq:svi-update})) to the running sufficient
  statistics, mixing the minibatch contribution with the prior
  pseudocounts and the previous EMA estimate.
\item Apply the closed-form M-steps of Section~\ref{sec:bw-tkf92}
  (quadratic for $\insrate,\delrate$, ratio for $\ext$, GTR coordinate
  ascent for $\exch,\eqm$).
\end{enumerate}

\paragraph{Comparison to SVI-BW.}
The two algorithms differ in their E-step and converge to different
fixed points:
\begin{itemize}[nosep]
\item SVI-BW (Section~\ref{sec:svi-bw}) treats each pair as
  conditionally independent given the population indel/substitution
  parameters, and sums over alignments with a 2D Pair-HMM
  Forward--Backward.  It uses no tree information.
\item svi-VarAnc takes a tree-structured MSA as input
  (alignment fixed, gap structure fixed, ancestral residues
  marginalised) and shares information across all branches in the
  tree at each MSA column.
\end{itemize}
For pairwise data (a tree of two leaves), the two algorithms coincide.
For deep trees with many siblings, svi-VarAnc tightens the indel-rate
posterior considerably by coupling sibling pairs through the tree.
ELBO monitoring, mini-batch construction, and warm-start
initialisation from cherry-trained TKF92 parameters follow the SVI-BW
recipe of Section~\ref{sec:svi-bw} verbatim, replacing the
Pair-HMM Forward--Backward E-step with the family-level VarAnc
sweep of Section~\ref{sec:felsenstein-update-tkf}.
