% implementations.tex — Appendix: Implementation Status & Validation
% Auto-maintained by Claude Code. Update when tests/modules change.
% Last updated: 2026-03-23

\section{Implementation Status and Validation}
\label{sec:appendix-implementations}

This appendix documents the JAX reference implementation of the algorithms
described in this paper, including module structure, CLI tools,
and the test suite used for validation.
The source code is in \texttt{python/tkfmixdom/jax/}.
All implementations use 64-bit floating point (via \texttt{jax.config.update("jax\_enable\_x64", True)}).
As of the current revision, the test suite contains 2099 tests (all passing).
The \texttt{vjp.py} module provides custom VJP wrappers for JAX-differentiable
log-likelihood computation, using Forward--Backward expected counts and analytic
score function derivatives in the backward pass.

\subsection{Package Overview}
\label{sec:impl-overview}

The \texttt{tkfmixdom} Python package (Python $\geq$ 3.10, JAX $\geq$ 0.4) implements
the full model hierarchy from TKF91 through the nested MixDom Pair HMM,
the grammar elaboration framework, order-1 distillation,
parse tree transducers, and beam-pruned phylogenetic inference.
Table~\ref{tab:impl-modules} summarizes the 58 modules.

\begin{table}[htbp]
\centering
\small
\caption{Implementation modules (\texttt{tkfmixdom.jax.*}).
Lines of code (LoC) are approximate.}
\label{tab:impl-modules}
\begin{tabular}{@{}llr@{}}
\toprule
\textbf{Module} & \textbf{Purpose} & \textbf{LoC} \\
\midrule
\multicolumn{3}{@{}l}{\emph{Core (\texttt{core/})}} \\
\texttt{params.py} & TKF91/92 transition matrices (joint \& conditioned) & 126 \\
\texttt{bdi.py} & BDI parameters, score derivatives, L'H\^opital limits, \texttt{bdi\_stats\_from\_counts} (eqs exposure-tkf--deaths-tkf), \texttt{m\_step\_indel\_quadratic} (eq kappa-quadratic) & 450 \\
\texttt{ctmc.py} & JC69 rate matrix, bridge expectations, CTMC prior & 160 \\
\texttt{ctmc\_irreversible.py} & General eigendecomp.\ matrix exp., irreversible bridge expectations & 156 \\
\texttt{protein.py} & LG substitution model, amino acid alphabets & 205 \\
\texttt{protein\_gap.py} & LG+gap model (21-state) with gap as extra character & 219 \\
\texttt{rna.py} & RNA 4-state alphabet, basepair definitions & 76 \\
\texttt{types.py} & Shared type definitions, constants & 241 \\
\midrule
\multicolumn{3}{@{}l}{\emph{Models (\texttt{models/})}} \\
\texttt{left\_regular.py} & TKF91/92 Pair HMMs as left-regular grammars & 140 \\
\texttt{mixdom.py} & Nested MixDom Pair HMM with null elimination & 408 \\
\texttt{elaborated.py} & Elaborated grammar model construction & 783 \\
\texttt{fully\_exploded.py} & Fully exploded MixDom (all null states explicit) & 446 \\
\texttt{chi\_free.py} & Autograd-able collapsed chi builder from free params (eqs~\ref{eq:main-counts}--\ref{eq:var-counts} verification) & 200 \\
\texttt{elimination\_steps.py} & 6-step null elimination chain, vectorized restoration & 124 \\
\texttt{exact\_suffstats.py} & Exact BDI suff.\ stats via chain restoration & 197 \\
\texttt{compiled.py} & Compiled model implementations for DP & 899 \\
\texttt{rna\_grammar.py} & RNA stem-loop SCFG, compound terminal encoding & 811 \\
\texttt{tkf\_grammar.py} & TKF grammar construction via elaboration rules & 584 \\
\texttt{context\_free.py} & TKFST pair SCFG (Section~7.3), IO, Rate EM, distillation & 2135 \\
\texttt{order1\_scfg.py} & Order-1 singlet SCFG (10 NTs, 260 rules) & 169 \\
\midrule
\multicolumn{3}{@{}l}{\emph{Grammar (\texttt{grammar/})}} \\
\texttt{scfg.py} & WCFG/SCFG: Inside/Outside/CYK, epsilon elimination & 786 \\
\texttt{scfg\_dp.py} & JAX-compiled Inside for WCFGs and Pair SCFGs & 199 \\
\texttt{compile.py} & Null removal/restoration for grammar compilation & 189 \\
\midrule
\multicolumn{3}{@{}l}{\emph{DP algorithms (\texttt{dp/})}} \\
\texttt{hmm.py} & Forward/Viterbi 1D \& 2D, padded FB, \texttt{safe\_log} & 1316 \\
\texttt{hmm\_beam.py} & Beam forward/backward for Pair HMMs & 563 \\
\texttt{scfg\_beam.py} & Phylogenetic Inside/Outside via WPTT pushing & 2390 \\
\texttt{scfg\_factored.py} & Factored Inside-Outside for SCFG $\times$ WPTT & 1681 \\
\midrule
\multicolumn{3}{@{}l}{\emph{Training (\texttt{train/})}} \\
\texttt{em\_tkf.py} & Baum--Welch EM for TKF models & 1028 \\
\texttt{em\_mixdom.py} & EM for nested MixDom model & 1207 \\
\texttt{em.py} & Generic EM framework & 188 \\
\texttt{vjp.py} & Custom VJP wrappers (joint \& conditioned) with log-posterior & 370 \\
\texttt{adam\_mixdom.py} & Adam optimizer for MixDom training & 416 \\
\texttt{likelihood.py} & Uniform differentiable LL API (observed + Q-function) & 210 \\
\texttt{constrained.py} & Constrained optimization utilities & 264 \\
\texttt{fit.py} & Pairwise TKF fitting pipeline & 167 \\
\texttt{optimizer.py} & Optimizer abstractions & 107 \\
\texttt{restricted\_mstep.py} & Rate-rescaling, tied-pi, and banded fragdist/ext M-step variants (sec.~\ref{sec:mstep-rescaling}, \ref{sec:mstep-tied-pi}) & 270 \\
\texttt{tkf92\_svi\_bw.py} & Pure-TKF92 SVI Baum--Welch (vmap'd batched 2D F--B over a bin-bucketed minibatch + per-pair-$t$ suff stats + closed-form $\kappa$-quadratic M-step) & 756 \\
\texttt{tkf92\_adam\_fb.py} & Pure-TKF92 Adam-on-F--B (vmap'd batched joint $(\lambda, \mu, \mathrm{ext})$ loss; GGI prior-swap option) & 649 \\
\midrule
\multicolumn{3}{@{}l}{\emph{Distillation (\texttt{distill/})}} \\
\texttt{hmm.py} & Order-1 HMM/WFST distillation & 473 \\
\texttt{maraschino.py} & Algebraic distillation via Woodbury identity & 713 \\
\texttt{maraschino\_fit.py} & MixDom cherry-count likelihood + raw $\leftrightarrow$ linear param conversion & 510 \\
\texttt{tkf92\_mixture.py} & K-component TKF92 mixture EM (Maraschino-style) & 700 \\
\texttt{banded\_mixture.py} & K-component banded MixDom mixture EM (Phase 2 of FragStart/FragMid/FragEnd pipeline; sec.~\ref{sec:mstep-tied-pi}) & 360 \\
\texttt{scfg.py} & Order-1 SCFG distillation with 4-position context & 813 \\
\texttt{wptt.py} & Weighted Parse Tree Transducer (246 states) & 424 \\
\midrule
\multicolumn{3}{@{}l}{\emph{Tree algorithms (\texttt{tree/})}} \\
\texttt{felsenstein.py} & Felsenstein pruning, tree likelihood & 169 \\
\texttt{ancestor.py} & Viterbi \& progressive ancestral reconstruction & 649 \\
\texttt{progrec\_felsenstein.py} & Profile-based ProgRec with order-1 WFST DP & 2163 \\
\texttt{progrec\_dag\_dp.py} & DAG-aware Viterbi/Forward for recognizer profiles & 729 \\
\texttt{progrec\_recognizer.py} & DAG recognizer for ProgRec (construction, compression) & 831 \\
\texttt{beam\_msa.py} & Beam-pruned MSA likelihood & 301 \\
\texttt{compose\_intersect.py} & WFST composition+intersection for ProgRec & 420 \\
\texttt{fsa\_anneal.py} & FSA sequence annealing (pairwise posteriors) & 600 \\
\texttt{constrained\_beam\_msa.py} & BeamMSA with ancestral gap constraints & 331 \\
\texttt{beam\_msa\_labeled.py} & Labeled BeamMSA with domain annotations & 350 \\
\texttt{progrec\_labeled.py} & Labeled ProgRec with domain annotations & 450 \\
\texttt{guide\_tree.py} & NJ, UPGMA, and star guide trees & 360 \\
\texttt{compose.py} & Transducer composition for trees & 257 \\
\texttt{compose\_wptt\_rec.py} & WPTT $\times$ Recognizer factored DP & 308 \\
\texttt{recognizer.py} & Leaf recognizer (span-based, consensus fallback) & 437 \\
\texttt{intersect.py} & Sibling recognizer intersection & 112 \\
\texttt{progressive.py} & Progressive reconstruction pipeline (SCFG) & 144 \\
\texttt{transducer.py} & Parse tree transducer, Pair SCFG Inside & 554 \\
\texttt{profile\_compress.py} & Profile SCFG compression (subtree sharing) & 612 \\
\texttt{partition\_recon.py} & Partition-conditioned reconstruction with $D{\times}F$ fragment tracking (ref) & 1388 \\
\texttt{partition\_recon\_jax.py} & Partition-conditioned reconstruction (JAX vectorised) & 459 \\
\midrule
\multicolumn{3}{@{}l}{\emph{Simulation (\texttt{simulate/})}} \\
\texttt{simulate.py} & Stochastic TKF91/92 sequence evolution & 374 \\
\texttt{evolve.py} & Pair HMM evolver via Forward traceback (plain TKF) & 174 \\
\texttt{pair\_hmm.py} & Pair HMM simulation utilities & 269 \\
\texttt{msa.py} & MSA generation from simulated evolution & 145 \\
\texttt{labeled\_utils.py} & Shared utilities for the MixDom fully-labeled simulators & 230 \\
\texttt{collapsed\_mixdom.py} & Simulator A: collapsed pair-HMM categorical sampler & 379 \\
\texttt{labeled\_mixdom.py} & Simulator B: hierarchical Gillespie with full sufficient-statistic tracking & 602 \\
\midrule
\multicolumn{3}{@{}l}{\emph{Evaluation (\texttt{evaluate/})}} \\
\texttt{metrics.py} & Centralized scoring: NW identity, SP/TC scores, annotation accuracy & 251 \\
\texttt{io.py} & Pipeline intermediate file I/O (NPZ+JSON, no data leakage) & 165 \\
\midrule
\multicolumn{3}{@{}l}{\emph{Utilities (\texttt{util/})}} \\
\texttt{io.py} & FASTA, Stockholm, Newick I/O & 324 \\
\texttt{data.py} & Pfam/Rfam/UniProt data download & 459 \\
\texttt{timing.py} & Timing utilities for benchmarks & 123 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{CLI Tools}
\label{sec:impl-cli}

Two command-line scripts (in \texttt{python/scripts/}) provide direct access
to sequence evolution:

\begin{description}[style=nextline,leftmargin=2em]
\item[\texttt{evolve.py}]
  Stochastic evolution under TKF91/TKF92 Pair HMMs via Forward traceback.
  Accepts ancestor sequences in FASTA format, produces descendant + alignment.
  Options: \texttt{-{}-model \{tkf91|tkf92\}}, \texttt{-{}-t}, \texttt{-{}-ins-rate},
  \texttt{-{}-del-rate}, \texttt{-{}-ext}, \texttt{-{}-n-samples}, \texttt{-{}-seed}.

\item[\texttt{evolve\_tree.py}]
  Parse tree transducer evolution: CYK parse of ancestor $\to$ stochastic tree
  transduction $\to$ descendant sequence.
  Additional options: \texttt{-{}-ins-prob}, \texttt{-{}-del-prob}, \texttt{-{}-show-tree}.

\item[\texttt{train\_pfam.py}]
  Train MixDom on Pfam via alignment-constrained Baum--Welch.
  Picks cherry pairs from Stockholm MSAs, runs 1D Forward--Backward
  with associative scan, accumulates exact BDI sufficient statistics
  via null count restoration.
  Features: pair manifest caching, streaming (constant memory),
  geometric-bin padded FB for JIT cache reuse, auto-checkpoint/resume,
  reversibility enforcement (both pair directions).
  Options: \texttt{-{}-msa-dir}, \texttt{-{}-n-dom}, \texttt{-{}-n-frag},
  \texttt{-{}-n-iter}, \texttt{-{}-checkpoint}, \texttt{-{}-families}.

\item[\texttt{maraschino.py}]
  CherryML-like distillation training. Four modes:
  \texttt{count} (extract adjacency counts from MSAs),
  \texttt{fit} (optimize MixDom params via Adam/L-BFGS),
  \texttt{distill} (generate order-1 Pair WFST / Singlet HMM),
  \texttt{fetch} (download Pfam Stockholm alignments).
\end{description}

\subsection{Key Algorithms and Their Implementation}
\label{sec:impl-algorithms}

\subsubsection{TKF91/92 Pair HMM DP}
The 2D Forward and Viterbi algorithms (\texttt{dp/hmm.py}) use an anti-diagonal wavefront
pattern suitable for \texttt{jax.lax.scan} parallelization:
cells on diagonal $d = i + j$ depend only on diagonals $d-1$ and $d-2$.
This yields $O(L)$ sequential steps with $O(L)$ parallel work per step.
Stochastic traceback (\texttt{sample\_traceback\_2d}) samples alignments
proportional to their Forward probability.

\subsubsection{MixDom Nested Pair HMM}
The nested TKF model (\texttt{models/mixdom.py}) constructs the
$7 \times 7$ exploded transition matrix with null states for between-domain,
between-fragment, and between-class transitions. Null elimination uses
$(I - T_{NN})^{-1}$ closure to produce the effective transition matrix.

\subsubsection{Grammar Elaboration}
The grammar framework (\texttt{grammar/scfg.py}, \texttt{models/tkf\_grammar.py}) implements
the seven elaboration rules of Appendix~\ref{sec:appendix-grammar}:
link, CTMC expansion, fragment, mixture, concatenation, non-recursive nesting,
and recursive nesting. Each rule transforms a WCFG by adding productions.
Epsilon elimination via nullability fixed-point and
$(I - W)^{-1}$ unary closure is provided for both singlet and pair grammars.

\subsubsection{Order-1 SCFG and WPTT}
The order-1 singlet SCFG (\texttt{models/order1\_scfg.py}) has 10 nonterminals
encoding 7-class basepair context (AU, CG, GC, UA, GU, UG, NN)
plus structural states (Start, L, R).
The Weighted Parse Tree Transducer (\texttt{distill/wptt.py}) has 246 states
($1 + 7 \times 7 \times 5$: Start plus all combinations of input context,
output context, and alignment type M/I/D/V/W).
SCFG structural weights can be folded into WPTT rule weights
(\texttt{scfg\_weighted\_wptt\_weights}) as an efficient approximation
to full SCFG $\times$ WPTT composition.

\subsubsection{Progressive Reconstruction (ProgRec)}
The ProgRec algorithm (\texttt{tree/progrec\_felsenstein.py}) performs
profile-based progressive alignment with Felsenstein conditional likelihoods.
At each internal node, two child profiles are aligned via pair HMM
Forward/Viterbi, optionally with stochastic sampling to build a
recognizer DAG (\texttt{tree/progrec\_recognizer.py}).
The recognizer DAG preserves alignment uncertainty: each path through
the DAG generates a valid subtree MSA. Compression merges
equivalent nodes (same wait-state connectivity, compatible leaf maps)
via bubble flattening. A NJ guide tree with Jukes-Cantor distance
correction provides the tree topology for benchmarking.

When using the MixDom model, ProgRec employs the full order-1 WFST
distilled from the domain mixture via algebraic distillation
(\texttt{distill/maraschino.py}).
The DP tracks 440 context states per cell:
$\text{AA} \times \text{AA} = 400$ for Match (left/right character context),
$\text{AA} = 20$ each for Insert and Delete contexts.
Transition probabilities ($p_{\text{mm}}, p_{\text{mi}}, p_{\text{md}},
p_{\text{im}}, \ldots$) depend on previously emitted characters,
capturing character-dependent correlations from the domain posterior.
Profile conditional likelihoods serve as Felsenstein emission factors.
Parent profiles are constructed using effective (domain-averaged)
substitution matrices.

\subsubsection{Partition-Conditioned Ancestral Reconstruction}
The partition-conditioned reconstruction algorithm
(\texttt{tree/partition\_recon.py}, reference Python;
\texttt{tree/partition\_recon\_jax.py}, JAX-vectorised) is formally
specified in Appendix~\ref{sec:partition-recon}.
It targets a restricted MixDom model in which the top-level TKF91
indel rates vanish ($\insrate_\main, \delrate_\main \to 0$ at fixed
ratio $\kappa_\main$) and each domain has a single fragment class.
For each starting column $k+1$ and each domain class $\dom$, the
algorithm maintains a per-branch TKF92 WFST state vector
$s_r(k+1, l)$ and incrementally accumulates
$\log \tilde G(k+1, l, \dom)$ as $l$ increases, reading off the
closed block likelihood $\log G(k+1, l, \dom)$ at every closing
position. Forward and Backward recursions over the block HMM combine
these block likelihoods with a geometric partition prior to produce
(i)~the total data log-likelihood, (ii)~a per-column posterior over
domain classes, and (iii)~a marginal posterior over root residues
obtained by mixing per-class Felsenstein root posteriors with the
per-column class posterior. The JAX implementation pads the MSA
length to the nearest geometric bin so that the compiled kernel is
reused across inputs. Both implementations agree to machine
precision on small inputs. An adapter
(\texttt{experiments/partition\_recon\_adapter.py}) exposes a call
signature compatible with the unified reconstruction benchmark, and
a driver (\texttt{experiments/partition\_recon\_benchmark.py}) runs
the method alongside the Felsenstein baseline on Pfam val families
without modifying the unified benchmark.

\subsubsection{Beam-Pruned Phylogenetic Inside/Outside}
The beam SCFG pipeline (\texttt{dp/scfg\_beam.py}) computes phylogenetic likelihood
of an MSA under the SCFG model via WPTT state pushing:
\begin{enumerate}
\item \textbf{Leaf CL}: For each leaf, compute $\mathrm{CL}_k[w, i, j]$
  via WPTT $\times$ Recognizer composition over MSA columns, handling gap
  columns with explicit delete transitions. ($w$: WPTT state; $[i,j)$: MSA span.)
\item \textbf{Internal nodes}: At bifurcation, children reset to Start:
  $\mathrm{CL}_B[\text{Start}, i, j] = \mathrm{CL}_L[\text{Start}, i, j] \cdot \mathrm{CL}_R[\text{Start}, i, j]$.
\item \textbf{Root}: $\log P(\text{MSA}) = \mathrm{CL}_{\text{root}}[\text{Start}, 0, L]$.
\item \textbf{Outside pass}: Reverse propagation from root span yields
  $O[w, r, i, j]$ for posterior computation.
\item \textbf{Posteriors}: Basepair probabilities
  $P(i \text{ paired with } j \mid \text{data})$ via LR-match transition
  posteriors; span posteriors $P([i,j) \text{ active} \mid \text{data})$.
\end{enumerate}
Beam pruning retains the top $B$ WPTT states per span.
CYK mode produces Viterbi parse with dot-bracket traceback.

\subsubsection{TKF Structure Tree (TKFST) Pair SCFG}
The TKFST grammar (\texttt{models/context\_free.py}) implements the full pair SCFG
from Section~7.3, with 49 nonterminals (M/I/D variants of each structural NT)
and $\sim$894 productions including lr-linear basepair rules.
The 4D Inside/Outside algorithms compute $I[A, i_x, j_x, i_y, j_y]$ and
$O[A, i_x, j_x, i_y, j_y]$ over ancestor and descendant span pairs.
Outside uses an augmented unary closure matrix that includes effective unary
transitions from binary rules with nullable children.
Rate EM classifies each rule by its parameter factor decomposition
($\alpha_S$, $\beta_L$, $\kappa_S$, etc.) and applies closed-form M-steps.
Distillation extracts stem basepair and loop singlet emission statistics
from expected counts.

\subsubsection{Custom VJP Wrappers}
All HMM-based models provide differentiable observed-data log-likelihoods
(\texttt{train/vjp.py}) using the score function identity:
$\partial(\log P)/\partial\theta = \sum_{ij} E[n_{ij}] \cdot \partial(\log \tau_{ij})/\partial\theta$.
The forward pass runs DP (expensive); the backward pass autodiffs
through transition matrix construction only (cheap).
Table~\ref{tab:impl-vjp} lists the available wrappers.

\begin{table}[htbp]
\centering\small
\caption{Custom VJP log-likelihood wrappers.}
\label{tab:impl-vjp}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Function} & \textbf{Model} & \textbf{Differentiable w.r.t.} & \textbf{Mode} \\
\midrule
\texttt{tkf91\_log\_prob} & TKF91 & $\lambda, \mu$ & Joint \\
\texttt{tkf91\_log\_prob\_cond} & TKF91 & $\lambda, \mu$ & Conditional \\
\texttt{tkf92\_log\_prob} & TKF92 & $\lambda, \mu, \ext^{(\dom)}_{fg}$ & Joint \\
\texttt{mixdom\_log\_prob} & MixDom & All structural params & Joint \\
\texttt{expected\_ll} (with subst) & All & ALL params incl.\ $Q$ & Q-function \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Baum--Welch EM for TKF91/92}
The TKF91/92 EM implementation (\texttt{train/em\_tkf.py}) alternates:
E-step via Forward--Backward expected counts,
substitution M-step via bridge-expectation sufficient statistics,
and indel M-step via closed-form BDI updates.
TKF92 decomposes the self-loop count into Markovian fragment transition
and new-fragment components using the $\nfrag \times \nfrag$ transition matrix
$\ext^{(\dom)}_{fg}$, then applies the TKF91 closed-form BDI update.
When log-likelihood deltas drop below a threshold,
L-BFGS fallback is available via custom VJP wrappers (\texttt{train/vjp.py}).

\subsubsection{Exact MixDom Baum--Welch via Chain Restoration}
The MixDom M-step is \emph{exact and closed-form} at every level via
the 6-step null elimination chain (\texttt{models/exact\_suffstats.py}).
The fully exploded MixDom Pair HMM (\texttt{models/fully\_exploded.py})
makes all null states explicit. The chain of 6 elimination steps
collapses this to the standard $5KF{+}2$ model; the chain of 6
restoration steps inverts this exactly.
Steps 1--5 have $C = I$ (trivial proportional splits).
Step 6 uses a $3 \times 3$ analytic DomEnd null closure.
The vectorized restoration (\texttt{models/elimination\_steps.py})
computes all DomEnd counts via matrix multiplications:
$\hat{n}_{K\silent} = T_{K\silent} \odot (\mathit{Scale} \cdot H^\top)$,
$\hat{n}_{\silent\silent} = T_{\silent\silent} \odot
(\widetilde{T}^\top \cdot \mathit{Scale} \cdot H^\top)$.
Verified against autodiff to $10^{-14}$ precision across
7 edge-case parameter configurations (14 tests in
\texttt{TestExactSuffStats}).
The production training script (\texttt{train\_pfam.py}) uses this
for all M-step updates.
Passing \texttt{None} for a callback freezes that parameter group.

\subsubsection{Fully-Labeled MixDom Simulators for Recovery Tests}
Two complementary simulators (\texttt{simulate/collapsed\_mixdom.py} and
\texttt{simulate/labeled\_mixdom.py}, with shared utilities in
\texttt{simulate/labeled\_utils.py}) drive the watertight parameter and
count recovery test suite for the per-pair-$t$ training pipeline.
\emph{Simulator A} (collapsed) walks the nested Pair HMM transition
matrix $\chi$ as a Markov chain at each pair's own branch length~$t$:
sample \texttt{next\_state} from $\chi[u, :]$, then sample the emitted
$(a, b)$ pair from the per-state per-class joint emission. No DP, just
two \texttt{np.random.choice} calls per transition; fast enough for
high-throughput parameter recovery on $B \in [400, 5000]$ batches.
\emph{Simulator B} (hierarchical Gillespie) generates events at every
level: a top-level TKF91 links process over domains, per-domain TKF92
links processes over fragments with sticky $\ext^{(\dom)}_{fg}$
extensions, per-fragment fragchar Markov chain, per-class CTMC
trajectory at each match site (recording $W_c$ dwell times and $U_c$
transition counts via accept-reject on the endpoint). Simulator B
tracks every complete-data sufficient statistic
($B$, $D$, $S$, $M$, $L$ at the top level and per domain;
$V_c$, $W_c$, $U_c$ per class) so that the EM E-step's expected
counts can be compared directly against ground truth. The single-fragchar special case is
recovered as the $n_\classes = 1$ special case. Both simulators
return the precompiled-pair tuple format consumed by
\texttt{train\_pfam.\_process\_pairs\_batched}, with explicit per-pair
$t$ — the central test condition for the per-pair-$t$ regression suite.

\subsection{Test Suite}
\label{sec:impl-tests}

The test suite (2099 tests, all passing) validates correctness
at multiple levels.
Tests are run with \texttt{python3.12 -m pytest tests/} from the
\texttt{python/} directory.
Table~\ref{tab:impl-tests} lists all test files with counts and descriptions.

\begin{table}[htbp]
\centering
\small
\caption{Test suite summary. Statistical tests are in \texttt{tests/statistical/}.}
\label{tab:impl-tests}
\begin{tabular}{@{}lrl@{}}
\toprule
\textbf{Test file} & \textbf{$n$} & \textbf{What is validated} \\
\midrule
\multicolumn{3}{@{}l}{\emph{Core model}} \\
\texttt{test\_params} & 9 & $\alpha, \beta, \gamma, \kappa$; row sums; TKF92$\to$91 \\
\texttt{test\_indel} & 42 & L'H\^opital limits; finite diff; Gillespie BDI; conservation law \\
\texttt{test\_subst} & 10 & JC69 rate matrix; $P(t)$ properties; bridge expectations \\
\texttt{test\_subst\_irreversible} & 17 & General eigendecomp.; irreversible bridge expectations; $\eqm$ recovery \\
\texttt{test\_dp} & 13 & Forward on identical/different; Fwd $\geq$ Viterbi \\
\texttt{test\_dp\_sampling} & 5 & Stochastic traceback validity and diversity \\
\texttt{test\_msa} & 4 & Pair emission; fixed-alignment likelihood \\
\texttt{test\_vjp} & 11 & Custom VJP vs.\ autodiff; TKF91/92 gradients \\
\texttt{test\_em} & 5 & E-step counts; EM iteration; likelihood progression \\
\texttt{test\_em\_mixdom} & 4 & MixDom EM training; parameter recovery \\
\texttt{test\_mixdom} & 5 & Nested Pair HMM; $n\_\text{states}(D,F)$; row sums \\
\texttt{test\_protein\_models} & 16 & WAG/LG rate matrices; reversibility \\
\texttt{test\_protein\_gap} & 12 & LG+gap model; gap-as-character vs.\ gap-as-missing \\
\midrule
\multicolumn{3}{@{}l}{\emph{Grammar \& elaboration}} \\
\texttt{test\_grammar} & 19 & WCFG construction; nullability; Inside/Outside/CYK \\
\texttt{test\_grammar\_lr} & 16 & Left-linear, LR-linear production types \\
\texttt{test\_grammar\_dp} & 4 & JAX Inside matches numpy; differentiability \\
\texttt{test\_tkf\_grammar} & 22 & TKF singlet/pair grammars; Inside $=$ Forward \\
\texttt{test\_elaborated} & 159 & Elaborated grammar; exact suff.\ stats; phantom events \\
\texttt{test\_compiled\_model} & 17 & Compiled model DP; forward/backward consistency \\
\texttt{test\_tree\_transducer} & 22 & Parse trees; CYK$\to$tree; Pair SCFG Inside \\
\texttt{test\_rna\_scfg} & 37 & RNA stem-loop SCFG; compound terminals \\
\texttt{test\_scfg\_distill} & 13 & Context-dependent distillation; beam Inside \\
\texttt{test\_distill} & 18 & Null closure; order-1 HMM distillation \\
\texttt{test\_tkfst\_grammar} & 92 & TKFST singlet/pair SCFG; Inside/Outside; EM; distillation \\
\texttt{test\_maraschino} & 53 & CherryML-like MixDom training; Woodbury distillation \\
\midrule
\multicolumn{3}{@{}l}{\emph{Parse tree transducers}} \\
\texttt{test\_rna\_context} & 12 & 7-class basepair context; canonical pairs \\
\texttt{test\_order1\_scfg} & 18 & 10-NT SCFG; production types; weights \\
\texttt{test\_wptt} & 23 & 246-state WPTT; rule types; TKF weights \\
\texttt{test\_recognizer} & 27 & Leaf recognizer; banding; consensus fallback \\
\texttt{test\_compose\_wptt\_rec} & 7 & WPTT $\times$ Recognizer composition \\
\texttt{test\_scfg\_compose} & 7 & SCFG $\times$ WPTT materialized composition \\
\texttt{test\_factored\_dp} & 64 & Factored SCFG $\times$ WPTT Inside-Outside; detailed counts \\
\texttt{test\_intersect} & 5 & Sibling recognizer intersection \\
\texttt{test\_progressive} & 7 & Progressive reconstruction pipeline \\
\midrule
\multicolumn{3}{@{}l}{\emph{Beam inference}} \\
\texttt{test\_beam} & 19 & Beam $\leq$ exact; wider $\to$ tighter; envelope \\
\texttt{test\_beam\_scfg} & 120 & Phylo Inside/CYK; CYK traceback; Outside; \\
& & posteriors; SCFG weights; cross-validation \\
\texttt{test\_beam\_msa} & 11 & Beam-pruned MSA likelihood; profile scoring \\
\midrule
\multicolumn{3}{@{}l}{\emph{Phylogenetics \& progressive alignment}} \\
\texttt{test\_tree} & 8 & Felsenstein pruning; gap-as-missing \\
\texttt{test\_rescaling} & 5 & Deep trees (20--50 taxa) without underflow \\
\texttt{test\_ancestor} & 14 & Viterbi/sampling reconstruction; tree traversal \\
\texttt{test\_partition\_recon} & 29 & Partition-conditioned reconstruction: forward $=$ backward; TKF92 cross-validation; Python $=$ JAX; $D{\times}F$ fragment tracking; intra-block backward $F{\cdot}B = G$; class-dependent emission; ext$>$0 consistency \\
\texttt{test\_inside\_guided} & 15 & Posterior ancestral; beam refinement \\
\texttt{test\_progrec\_felsenstein} & 25 & ProgRec profile alignment; MSA extraction; recognizer sampling \\
\texttt{test\_progrec\_mixdom} & 21 & Order-1 WFST distillation; context-dependent DP; MixDom ProgRec \\
\texttt{test\_progrec\_dag\_dp} & 18 & DAG-aware Viterbi/Forward; leaf DAG; MSA extraction \\
\texttt{test\_progrec\_recognizer} & 29 & DAG recognizer; construction; compression; MSA invariant \\
\texttt{test\_guide\_tree} & 13 & TKF92 distance estimation; NJ tree; guide tree pipeline \\
\texttt{test\_evolve} & 10 & Stochastic evolution; TKF92$\to$91; length dist.\ \\
\texttt{test\_io} & 15 & FASTA/Stockholm/Newick roundtrip \\
\texttt{test\_data} & 4 & Sequence encoding; data fetch \\
\texttt{test\_fit} & 7 & Pairwise fitting; branch length estimation \\
\texttt{test\_integration} & 7 & Simulate $\to$ Fit $\to$ Reconstruct roundtrip \\
\texttt{test\_cross\_validation} & 15 & Grammar $=$ Forward; JAX $=$ numpy \\
\texttt{test\_benchmark} & 15 & Benchmark framework; performance regression \\
\midrule
\multicolumn{3}{@{}l}{\emph{Training validation}} \\
\texttt{test\_bw\_vs\_grad} & 22 & Baum--Welch vs.\ gradient consistency \\
\texttt{test\_grad\_consistency} & 18 & Gradient correctness across parameterizations \\
\texttt{test\_bdi\_new} & 28 & \texttt{bdi\_stats\_from\_counts} conservation/non-negativity/L'H\^opital; \texttt{m\_step\_indel\_quadratic} validity/LL improvement/NaN regression; Q\_exploded+prior monotonicity per parameter group across 4 edge-case configs; autograd count verification; aggregate EM \\
\texttt{test\_tkf92\_svi\_bw\_batched} & 7 & vmap'd batched E-step equivalence vs.\ independent pre-refactor numpy reference; padding-invariance (Adam-fix analogue); forward-only val\_eval; Q-function gradient zero at M-step solution (catches L/M coefficient mis-grouping) \\
\midrule
\multicolumn{3}{@{}l}{\emph{Statistical validation (\texttt{statistical/})}} \\
\texttt{test\_bdi\_stats} & 16 & Conservation law $E[B]-E[D]=j-i$; score derivs. \\
\texttt{test\_general\_bdi} & 35 & General BDI ($\nu$ free); reduces to TKF91 \\
\texttt{test\_baum\_welch} & 14 & EM monotonicity; parameter convergence; L-BFGS fallback \\
\texttt{test\_alignment\_recovery} & 5 & Viterbi alignment accuracy (sens., PPV) \\
\texttt{test\_holmes\_rubin} & 8 & Substitution expected dwell/transition counts \\
\texttt{test\_tree\_discrimination} & 3 & Phylogenetic signal detection \\
\texttt{test\_estep\_vs\_simulation} & 11 & E-step expected counts match simulation \\
\texttt{test\_mixdom\_simulation\_recovery} & 14 & Per-pair-$t$ parameter and count recovery via Simulators A and B (single- and multi-fragchar; mixed-$t$ batches); BDI/HR conversion $t$-sensitivity \\
\midrule
\textbf{Total} & \textbf{1693} & \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Testing Patterns}

The following testing strategies are employed throughout:

\begin{description}[style=nextline,leftmargin=2em]
\item[Cross-validation]
  Multiple implementations of the same quantity are compared:
  grammar Inside vs.\ Forward DP; JAX-compiled vs.\ numpy;
  beam SCFG CL vs.\ factored Inside; TKF92(ext$=$0) $\equiv$ TKF91.

\item[Probability invariants]
  Transition matrices have row sums $= 1$; grammar marginals $\leq 1$;
  posterior probabilities in $[0, 1]$; Inside $\geq$ Viterbi (logsumexp $\geq$ max).

\item[Statistical consistency]
  Simulated BDI counts converge to theoretical expectations
  with $\sqrt{N}$ tolerance.
  Conservation law $E[B] - E[D] = j - i$ validated via autodiff.
  Alignment recovery accuracy measured across parameter regimes.

\item[Monotonicity]
  EM likelihood non-decreasing; wider beam $\to$ tighter probability bound;
  wider band $\to$ better Inside approximation.

\item[Roundtrip integrity]
  Simulate $\to$ Fit $\to$ Reconstruct: recovered parameters close to truth.
  FASTA/Stockholm/Newick I/O roundtrip preserves content.
  Grammar $\to$ HMM conversion preserves transition structure.

\item[Edge cases]
  Empty sequences, single nucleotides, identical sequences, all-gap columns,
  deep trees (50 taxa), degenerate parameters ($t = 0$, $t \to \infty$).

\item[Finite-difference validation]
  Analytic score derivatives match numerical $\Delta f / \Delta\theta$
  across multiple parameter regimes.
\end{description}

\subsection{Benchmark Infrastructure}
\label{sec:impl-benchmarks}

Architecture-tagged benchmarks (\texttt{python/benchmarks/}) measure
wall-clock time for Forward, Viterbi, beam forward, grammar Inside
(numpy and JAX), progressive reconstruction, and Pair SCFG Inside
across sequence lengths $L \in \{10, 20, 50, 100, 200\}$.
Results are organized by hostname/CPU model and compiled into
a \LaTeX\ report with tables and plots via \texttt{make} (see
\texttt{benchmarks/Makefile}).

MSA accuracy benchmarks (\texttt{maraschino-paper/benchmarks/}) evaluate
ProgRec against MAFFT, MUSCLE, PRANK, and Historian on BAliBASE,
PREFAB, OXBench, and SABmark using SP (sum-of-pairs) and TC (total column)
scores. Variants include Viterbi-only and recognizer sampling with
NJ or balanced guide trees.
