\section{Models and Fitting Pipelines}
\label{sec:models-fitted}

We evaluate a hierarchy of substitution-plus-indel models,
ranging from a single reversible substitution matrix (Felsenstein)
to the full Nested TKF model with per-domain substitution profiles.
Each model is trained on the Pfam seed alignment database
and evaluated on a held-out test set (Section~\ref{sec:sim-eval}).

\subsection{Model hierarchy}

\begin{center}
\begin{tabular}{llccp{5cm}}
\textbf{ID} & \textbf{Description} & $|\ndom|$ & $\symsub$ & \textbf{Order-1 source} \\
\hline
F & Felsenstein/LG08 & --- & LG08 & None (order-0) \\
M3 & MixDom & 3 & shared & Domain structure \\
M6 & MixDom & 6 & shared & Domain structure \\
M8s & MixDom, SVI-BW & 8 & per-domain & Domain structure + per-domain $\symsub$ \\
\end{tabular}
\end{center}

All MixDom models include the TKF91 top-level indel process
and TKF92-like per-domain fragment extension.
Each domain type has its own equilibrium distribution $\eqm_\dom$
and (optionally) exchangeability $\symsub_\dom$
(Section~\ref{sec:per-domain-subst}).

\subsection{Felsenstein baseline (F)}

The simplest model: a single reversible substitution matrix (LG08~\cite{LG08})
applied independently to each column, with no indel model and no
inter-column correlations.
Ancestral reconstruction uses standard Felsenstein pruning.
With discretized gamma rate heterogeneity ($G = 4$ categories),
this becomes the LG08+$\Gamma$ model.

No training is required; the LG08 matrix and gamma shape parameter
($\alpha$) are taken from the literature.

\subsection{MixDom (M3, M6)}
\label{sec:mixdom-static-fitting}

The Nested TKF model with $|\ndom|$ domain types, each with
a single fragment type and shared exchangeability $\symsub$ (LG08).
Per-domain parameters: insertion rate $\insrate_\dom$,
deletion rate $\delrate_\dom$, domain weight $v_\dom$,
and fragment extension rate $r_\dom$.

\subsubsection{Stage 1: Maraschino (cherry-count fitting)}

Adjacent (cherry) pairs are extracted from Pfam seed alignments
and binned by estimated divergence time $\tau$ into 32 geometric bins.
For each bin, pairwise adjacency counts are accumulated:
match-match, match-insert, match-delete, etc.\
(see Section~\ref{sec:models-fitted} for the count types).

The MixDom parameters are fitted by maximizing the cherry-count
log-likelihood via Adam~\cite{Adam} (3000 steps, learning rate $10^{-3}$).
The cherry-count likelihood marginalizes the pair HMM over
alignment paths, using the algebraic distillation
(Section~\ref{sec:distill-hmm}) to compute order-1 adjacency frequencies
without enumerating alignments.

This stage fits: $\insrate_0, \delrate_0$ (top-level),
$\insrate_\dom, \delrate_\dom, v_\dom, r_\dom$ (per-domain),
$\symsub$ (shared exchangeability, initialized from LG08),
and $\alpha$ (gamma shape).

\subsubsection{Stage 2: SVI Baum--Welch refinement}

The Maraschino-fitted parameters are refined using
stochastic variational inference Baum--Welch (SVI-BW)
on the full pair HMM.
Mini-batches of sequence pairs (batch size 200) are drawn from
Pfam seed alignments; the E-step runs constrained 1D forward-backward
on each pair, and the M-step updates parameters from
the accumulated sufficient statistics.

SVI-BW refines the indel rates, domain weights, and
(with \texttt{--estimate-subst}) the per-domain substitution model
$(Q_\dom, \eqm_\dom)$ via bridge expectations.
The step size decays as $\eta_k = (k + \tau)^{-\kappa}$
with $\kappa = 0.6$, $\tau = 10$.

Model M8s uses SVI-BW with \texttt{--estimate-subst} to obtain
8 independently estimated per-domain exchangeability matrices
$\symsub^{(\dom)}$, giving each domain its own substitution profile.

\subsection{Distillation to order-1 machines}
\label{sec:distillation-pipeline}

All MixDom models are distilled to order-1 singlet HMMs and pair WFSTs
for use in TreeVarAnc and alignment.
The distillation uses the Woodbury identity
(Section~\ref{sec:distill-hmm}) to compute adjacency frequencies
$f^{\mat\mat}$, $f^{\mat\ins}$, etc., from which the order-1
transition probabilities are derived.

Each domain contributes its own per-domain emission tensor
$e_{\mat}(\anctok, \destok) = \eqm^{(\dom)}_\anctok P_\dom(\evoltime)_{\anctok,\destok}$
to the adjacency sums, with structural weights from the Woodbury decomposition.

The distilled WFSTs are cached by geometric time bin
(20 bins from $t = 0.001$ to $t = 8.0$) to avoid redundant
computation across branches with similar lengths.

\subsection{Domain sweep}
\label{sec:domain-sweep}

To select the optimal number of domains $|\ndom|$, we performed
a sweep over $|\ndom| \in \{3, 4, 5, 6, 8, 10, 12\}$.
For each $|\ndom|$, a MixDom model was fitted via Maraschino
(2000 Adam steps) on the training split of Pfam seed counts,
then evaluated on the held-out test split.

The test-set cherry-count log-likelihood peaked at $|\ndom| = 6$,
with $|\ndom| = 12$ as runner-up ($\Delta = 3117$ nats).
Models with $|\ndom| > 6$ showed signs of overfitting:
improving train LL but degrading test LL.
We use $|\ndom| = 6$ as the default for subsequent experiments.
