% tkf92-wfst-derivation.tex — TKF92 conditional WFST by singlet division
% To be \input from tkf.tex

\section{TKF92 WFST by Singlet Division}
\label{sec:tkf92-wfst}

This appendix derives the TKF92 conditional WFST
$\tkfwfst'(\insrate,\delrate,\evoltime,\ext)$
representing $P(\text{descendant} \mid \text{ancestor}, \theta)$
by dividing the TKF92 Pair HMM by the TKF92 Singlet HMM.

The TKF91 WFST (Section~\ref{sec:tkf91-machines}) is a $5{\times}5$
matrix obtained from the TKF91 Pair HMM $\tkftrans$ by stripping
$\kappa$ from ancestor-consuming columns ($\mat,\del$) and $1{-}\kappa$
from the end column ($\fin$).
The TKF92 WFST cannot be expressed as a $5{\times}5$ matrix in the same
state space.  As we show in Section~\ref{sec:wfst92-i0i1-split},
the singlet's outgoing-emission weight differs between its
$\sta$ state and its emit state, so the two cases must be carried as
distinct insert states (\emph{Ins0} and \emph{Ins1}) in the WFST.
The natural object is therefore $6{\times}6$, with state space
$\{\sta,\mat,\ins_0,\ins_1,\del,\fin\}$.
A $5{\times}5$ presentation that lumps $\ins_0$ and $\ins_1$ into a
single $\ins$ state and uses the emit-state divisor $p$ throughout is
incorrect for paths that traverse $\sta\to\ins\to\cdots\to\{\mat,\del\}$;
this is discussed in Section~\ref{sec:wfst92-vs-5x5}.

\subsection{TKF92 Singlet HMM Transitions}

The TKF92 singlet at stationarity emits sequences whose length $L$
follows a zero-inflated geometric distribution
(Section~\ref{sec:tkf92-machines}).
The singlet HMM has two transition contexts:

\begin{itemize}[nosep]
\item \emph{From $\sta$ (before any character):}
  $P(\text{emit first character}) = \kappa$,\quad
  $P(\text{end}) = 1-\kappa$.

\item \emph{From an emitting state (after character $i \geq 1$):}
  $P(\text{emit next character}) = p \equiv \ext + (1{-}\ext)\kappa$,\quad
  $P(\text{end}) = (1{-}\ext)(1{-}\kappa)$.
\end{itemize}
The effective continuation probability $p$ decomposes as:
fragment extension with probability $\ext$,
or fragment termination ($1{-}\ext$) followed by a new nonempty link ($\kappa$).
Each emitted character is drawn i.i.d.\ from $\eqm$.

The key consequence for the WFST construction below is that
$\kappa \neq p$ whenever $\ext > 0$.

\subsection{Why an Ins0 / Ins1 Split is Necessary}
\label{sec:wfst92-i0i1-split}

Transducer composition $\text{Singlet} \circ \text{WFST} = \text{Pair HMM}$
couples each ancestor-consuming Pair HMM transition to one of the two
singlet emission events:
\begin{itemize}[nosep]
\item the $\sta\to\text{emit}$ event (weight $\kappa$), used \emph{exactly once},
  for the first ancestor character; or
\item the $\text{emit}\to\text{emit}$ event (weight $p$), used for every
  subsequent ancestor character.
\end{itemize}
The Pair HMM's $\ins$ state corresponds to ``the descendant is in the
middle of a descendant-only fragment.''  This state is reached either
\emph{before} the first ancestor character has been consumed (entered
from $\sta$, possibly via $\ins\to\ins$ self-loops) or \emph{after}
$\geq 1$ ancestor characters have been consumed (entered from
$\mat,\del,\ins$).  In the first case, the next ancestor-consuming
transition out of $\ins$ corresponds to the $\sta\to\text{emit}$
event in the singlet (weight $\kappa$); in the second case, it
corresponds to $\text{emit}\to\text{emit}$ (weight $p$).
The two divisors differ unless $\ext=0$.

The $5{\times}5$ Pair HMM
$\tkftrans'(\insrate,\delrate,\evoltime,\ext)$ of
Section~\ref{sec:tkf92-machines} marginalises over the
``has any ancestor character been consumed yet?'' bit and is correct
as a generator of the joint distribution.  But because that bit
controls which singlet divisor applies, the WFST construction must
expose it.  We therefore split the Pair HMM $\ins$ state into
\begin{itemize}[nosep]
\item $\ins_0$ --- reachable only from $\sta$ (via $\sta\to\ins_0$) and from itself
  (via $\ins_0\to\ins_0$): the singlet has not yet emitted.
\item $\ins_1$ --- reachable from $\mat,\del,\ins_1$: the singlet has emitted
  at least one character.
\end{itemize}
The $6{\times}6$ Pair HMM in this state space is row-stochastic
(verified below), and the $5{\times}5$ Pair HMM is recovered by lumping
$\ins_0$ and $\ins_1$ into a single $\ins$ row.  The lumping preserves
the joint distribution because the outgoing transitions from $\ins_0$
and $\ins_1$ have identical Pair HMM structure once weighted by their
singlet contributions; what fails to lump is the WFST.

\subsection{The \texorpdfstring{$6{\times}6$}{6x6} TKF92 Pair HMM}
\label{sec:wfst92-pair-hmm}

Writing $\alpha,\beta,\gamma,\kappa$ for the TKF link-level parameters
(as in Section~\ref{sec:tkf91-machines}), and $r=\ext$
for the fragment-extension probability:
\begin{eqnarray}
\lefteqn{\tkftrans''(\insrate,\delrate,\evoltime,\ext) \;= \quad} \label{eq:tkf92-pair-hmm-6x6} \\
& & \renewcommand{\arraystretch}{1.6}
\left(
\begin{array}{r|cccccc}
& \sta & \mat & \ins_0 & \ins_1 & \del & \fin \\\hline
\sta   & 0 & (1{-}\beta)\kappa\alpha
            & \beta
            & 0
            & (1{-}\beta)\kappa(1{-}\alpha)
            & (1{-}\beta)(1{-}\kappa) \\
\mat   & 0 & r + (1{-}r)(1{-}\beta)\kappa\alpha
            & 0
            & (1{-}r)\beta
            & (1{-}r)(1{-}\beta)\kappa(1{-}\alpha)
            & (1{-}r)(1{-}\beta)(1{-}\kappa) \\
\ins_0 & 0 & (1{-}r)(1{-}\beta)\kappa\alpha
            & r + (1{-}r)\beta
            & 0
            & (1{-}r)(1{-}\beta)\kappa(1{-}\alpha)
            & (1{-}r)(1{-}\beta)(1{-}\kappa) \\
\ins_1 & 0 & (1{-}r)(1{-}\beta)\kappa\alpha
            & 0
            & r + (1{-}r)\beta
            & (1{-}r)(1{-}\beta)\kappa(1{-}\alpha)
            & (1{-}r)(1{-}\beta)(1{-}\kappa) \\
\del   & 0 & (1{-}r)(1{-}\gamma)\kappa\alpha
            & 0
            & (1{-}r)\gamma
            & r + (1{-}r)(1{-}\gamma)\kappa(1{-}\alpha)
            & (1{-}r)(1{-}\gamma)(1{-}\kappa) \\
\fin   & 0 & 0 & 0 & 0 & 0 & 0
\end{array}
\right) \nonumber
\end{eqnarray}
The structural zeros are: $\sta$ cannot reach $\ins_1$ in one step
(no ancestor has yet been emitted), and $\mat,\del,\ins_1$ cannot
reach $\ins_0$ in one step (an ancestor has already been emitted).
The $\ins_0$ and $\ins_1$ rows differ only in their self-loop column
($\ins_0\to\ins_0$ vs $\ins_1\to\ins_1$); their entries to $\mat,\del,
\fin$ are identical.

\paragraph{Row-stochasticity.}
For each non-$\fin$ source row, the row sum equals
\[
r + (1{-}r)\Bigl[\beta + (1{-}\beta)\kappa\alpha + (1{-}\beta)\kappa(1{-}\alpha)
                  + (1{-}\beta)(1{-}\kappa)\Bigr]
= r + (1{-}r)\bigl[\beta + (1{-}\beta)\bigr] = 1
\]
(replacing $\beta$ with $\gamma$ for the $\del$ row), confirming that
$\tkftrans''$ is a valid Markov transition matrix.

\paragraph{Recovery of the $5{\times}5$ Pair HMM.}
Lumping $\ins_0$ and $\ins_1$ into a single $\ins$ state preserves
all four non-$\ins$-column entries (which are identical between the
two rows by construction).  In the self-loop column,
$\ins_0\to\ins_0$ and $\ins_1\to\ins_1$ both equal $r+(1{-}r)\beta$,
so the lumped self-loop is also $r+(1{-}r)\beta$.  The result is the
$5{\times}5$ Pair HMM of Section~\ref{sec:tkf92-machines}, recovered
exactly.

\subsection{The \texorpdfstring{$6{\times}6$}{6x6} TKF92 WFST}
\label{sec:wfst92-6x6}

Each Pair HMM transition consumes either one ancestor character
($\to\mat$ or $\to\del$), zero ancestor characters
($\to\ins_0$ or $\to\ins_1$), or signals termination ($\to\fin$).
The WFST $\tkfwfst''$ is obtained by dividing each Pair HMM
entry by the corresponding singlet weight:

\begin{itemize}[nosep]
\item Ancestor-consuming entries from $\sta,\ins_0$
  (the next ancestor character is the \emph{first}):
  divide by $\kappa$.
\item Ancestor-consuming entries from $\mat,\ins_1,\del$
  (the next ancestor character is a \emph{subsequent} one):
  divide by $p = \ext + (1{-}\ext)\kappa$.
\item $\fin$ entries from $\sta,\ins_0$
  (the ancestor sequence has length zero):
  divide by $1{-}\kappa$.
\item $\fin$ entries from $\mat,\ins_1,\del$
  (the ancestor sequence has length $\geq 1$):
  divide by $(1{-}\ext)(1{-}\kappa)$.
\item $\ins_0$ and $\ins_1$ entries (no ancestor consumed): divide by $1$.
\end{itemize}

The emission factor $\eqm_\anctok$ is also stripped from the emission
weights, exactly as in TKF91, so that $\mat$-state emission becomes
$\exp(\revsub\evoltime)_{\anctok\destok}$ and $\del$-state emission
becomes $1$; we focus on the transition matrix below.

\begin{eqnarray}
\lefteqn{\tkfwfst''(\insrate,\delrate,\evoltime,\ext) \;= \quad} \label{eq:tkf92-wfst} \\
& & \renewcommand{\arraystretch}{1.7}
\left(
\begin{array}{r|cccccc}
& \sta & \mat & \ins_0 & \ins_1 & \del & \fin \\\hline
\sta   & 0 & (1{-}\beta)\alpha
            & \beta
            & 0
            & (1{-}\beta)(1{-}\alpha)
            & 1{-}\beta \\
\mat   & 0 & \dfrac{r + (1{-}r)(1{-}\beta)\kappa\alpha}{p}
            & 0
            & (1{-}r)\beta
            & \dfrac{(1{-}r)(1{-}\beta)\kappa(1{-}\alpha)}{p}
            & 1{-}\beta \\
\ins_0 & 0 & (1{-}r)(1{-}\beta)\alpha
            & r + (1{-}r)\beta
            & 0
            & (1{-}r)(1{-}\beta)(1{-}\alpha)
            & (1{-}r)(1{-}\beta) \\
\ins_1 & 0 & \dfrac{(1{-}r)(1{-}\beta)\kappa\alpha}{p}
            & 0
            & r + (1{-}r)\beta
            & \dfrac{(1{-}r)(1{-}\beta)\kappa(1{-}\alpha)}{p}
            & 1{-}\beta \\
\del   & 0 & \dfrac{(1{-}r)(1{-}\gamma)\kappa\alpha}{p}
            & 0
            & (1{-}r)\gamma
            & \dfrac{r + (1{-}r)(1{-}\gamma)\kappa(1{-}\alpha)}{p}
            & 1{-}\gamma \\
\fin   & 0 & 0 & 0 & 0 & 0 & 0
\end{array}
\right) \nonumber
\end{eqnarray}

The $\ins_1$ row uses the $p$ divisor on its $\mat$ and $\del$ exits
and $1{-}\beta$ on its $\fin$ exit, matching the post-immortal-link
regime.  The $\ins_0$ row uses $\kappa$ on its $\mat,\del$ exits
(absorbed into the numerator: $(1{-}r)(1{-}\beta)\kappa\alpha/\kappa
= (1{-}r)(1{-}\beta)\alpha$, etc.) and $1{-}\kappa$ on its $\fin$
exit (giving $(1{-}r)(1{-}\beta)$).  These are exactly the TKF91
WFST $\sta/\mat/\ins$-row outputs scaled by $(1{-}r)$, reflecting
the leading-extension fragment's chance of terminating before any
ancestor is consumed.

\subsection{Structure and Verification}
\label{sec:wfst92-verification}

\paragraph{Sanity check 1: $\sta$ row.}
The $\sta$ row inherits the TKF91 form because the first character
of the immortal-link descendant fragment has no extension self-loop.
Multiplying by the singlet emission weights from $\sta$
($\kappa$ for $\mat,\del$, $1$ for $\ins_0$, $1{-}\kappa$ for $\fin$)
recovers the Pair HMM $\sta$ row of \eqref{eq:tkf92-pair-hmm-6x6}.

\paragraph{Sanity check 2: $\ins_0$ row.}
Multiplying $\tkfwfst''$ entries by the appropriate singlet weights
(``from~$\sta$''~weights: $\kappa$ for $\mat/\del$, $1$ for $\ins_0$,
$1{-}\kappa$ for $\fin$):
\begin{align*}
\kappa\cdot(1{-}r)(1{-}\beta)\alpha &= (1{-}r)(1{-}\beta)\kappa\alpha
\quad =\;\tkftrans''_{\ins_0\mat},\\
1\cdot[r+(1{-}r)\beta] &= r+(1{-}r)\beta
\quad =\;\tkftrans''_{\ins_0\ins_0},\\
\kappa\cdot(1{-}r)(1{-}\beta)(1{-}\alpha) &= (1{-}r)(1{-}\beta)\kappa(1{-}\alpha)
\quad =\;\tkftrans''_{\ins_0\del},\\
(1{-}\kappa)\cdot(1{-}r)(1{-}\beta) &= (1{-}r)(1{-}\beta)(1{-}\kappa)
\quad =\;\tkftrans''_{\ins_0\fin}.
\end{align*}
All four entries match the Pair HMM, so $\text{Singlet}\circ\tkfwfst''=
\tkftrans''$ on the $\ins_0$ row.

\paragraph{Sanity check 3: $\mat$ row.}
Multiplying by emit-state weights ($p$ for $\mat/\del$, $1$ for
$\ins_1$, $(1{-}\ext)(1{-}\kappa)$ for $\fin$):
\begin{align*}
p\cdot\frac{r+(1{-}r)(1{-}\beta)\kappa\alpha}{p}
   &= r+(1{-}r)(1{-}\beta)\kappa\alpha
   \;=\;\tkftrans''_{\mat\mat},\\
1\cdot(1{-}r)\beta &= (1{-}r)\beta
   \;=\;\tkftrans''_{\mat\ins_1},\\
p\cdot\frac{(1{-}r)(1{-}\beta)\kappa(1{-}\alpha)}{p}
   &= (1{-}r)(1{-}\beta)\kappa(1{-}\alpha)
   \;=\;\tkftrans''_{\mat\del},\\
(1{-}r)(1{-}\kappa)\cdot(1{-}\beta) &= (1{-}r)(1{-}\beta)(1{-}\kappa)
   \;=\;\tkftrans''_{\mat\fin}.
\end{align*}

The $\ins_1$ and $\del$ rows verify identically (with $\gamma$ for
$\beta$ in the $\del$ row).

\paragraph{TKF91 limit ($\ext=0$).}
At $\ext=0$, $p=\kappa$ and $r=0$.  The $\ins_0$ and $\ins_1$ rows
become:
\begin{align*}
\tkfwfst''_{\ins_0\,*\;|\,r=0}
  &= \bigl(0,\;(1{-}\beta)\alpha,\;\beta,\;0,\;(1{-}\beta)(1{-}\alpha),\;1{-}\beta\bigr),\\
\tkfwfst''_{\ins_1\,*\;|\,r=0}
  &= \bigl(0,\;(1{-}\beta)\alpha,\;0,\;\beta,\;(1{-}\beta)(1{-}\alpha),\;1{-}\beta\bigr).
\end{align*}
With the $\ins_0$ and $\ins_1$ self-loops both equal to $\beta$ and
all other columns identical, the two rows reduce to the same
TKF91-Ins row when $\ins_0,\ins_1$ are merged.  The TKF91 WFST
(Section~\ref{sec:tkf91-machines}) is recovered.

\paragraph{Latent fragment-boundary posterior.}
The $\mat$ row's $\mat\to\mat$ entry decomposes as
\[
\tkfwfst''_{\mat\mat}
= \underbrace{\frac{r}{p}}_{\substack{\text{same}\\\text{frag}}} \cdot 1
\;+\;
\underbrace{\frac{(1{-}r)\kappa}{p}}_{\substack{\text{new}\\\text{frag}}} \cdot (1{-}\beta)\alpha,
\]
exactly as in the original construction; under same-fragment
continuation the descendant certainly stays in $\mat$
(the link persists within the fragment), and under new-fragment
start the descendant must survive ($1{-}\beta$) and align ($\alpha$).
The fragment-boundary posterior $r/p + (1{-}r)\kappa/p = 1$ holds by
definition of $p$.  The $\ins_0$ row has \emph{no} same-fragment
ancestor-consuming term --- since the singlet has not yet emitted,
the fragment-boundary posterior is degenerate at ``new fragment'' ---
which is why the $\ins_0\to\mat$ entry has the simple
$(1{-}r)(1{-}\beta)\alpha$ form rather than a $1/p$-divided form.

\subsection{Comparison with the \texorpdfstring{$5{\times}5$}{5x5} Form}
\label{sec:wfst92-vs-5x5}

A $5{\times}5$ presentation that replaces the two rows
$\ins_0,\ins_1$ of \eqref{eq:tkf92-wfst} by a single $\ins$ row equal
to the $\ins_1$ row is correct only at $\ext=0$.  For $\ext>0$, that
$5{\times}5$ form misweights paths
$\sta \to \ins \to \cdots \to \{\mat,\del,\fin\}$ with one or more
leading inserts before the first ancestor consumption: the divisor
on the leading-insert exit should be $\kappa$ (or $1{-}\kappa$ for
$\fin$) rather than $p$ (or $(1{-}\ext)(1{-}\kappa)$).
For a single-leading-insert path $\sta\to\ins\to\mat\to\fin$,
the multiplicative discrepancy is
$(1{-}\ext)\kappa/p < 1$ when $\ext>0$, so the $5{\times}5$ WFST
\emph{underweights} such paths.

For benchmarks dominated by alignments without leading inserts
(typical when the ancestor's $N$-terminus is well-conserved),
the numerical effect of the discrepancy is small per pair.
However, the $5{\times}5$ form does not represent the true
conditional distribution $P(\text{descendant} \mid \text{ancestor},
\theta)$, and the $6{\times}6$ WFST in \eqref{eq:tkf92-wfst} is
the construction we recommend.

\subsection{Normalization Structure}
\label{sec:wfst-row-sums}

The WFST is conditionally normalised in the global sense: for every
input sequence $x \in \Sigma^*$,
\[
\sum_{(y,\,\text{alignment})}
   \bigl[\text{WFST path-weight}\bigr]
\;=\; 1,
\]
where the sum is over all paths from $\sta$ to $\fin$ that read $x$
exactly.  This is a direct consequence of
$\text{Singlet} \circ \tkfwfst'' = \tkftrans''$ (verified row by row
in Section~\ref{sec:wfst92-verification}) combined with the fact that
$\tkftrans''$ is row-stochastic and the Singlet HMM defines a proper
distribution on $x$.

The WFST is \emph{not} stochastic in the local single-step sense:
its rows do not sum to $1$, and the divisor structure
(ancestor-consuming columns divide by $\kappa$ from $\sta,\ins_0$ and
by $p$ from $\mat,\ins_1,\del$; the $\fin$ column divides by $1{-}\kappa$
from $\sta,\ins_0$ and by $(1{-}\ext)(1{-}\kappa)$ from
$\mat,\ins_1,\del$; insert columns divide by $1$) means a fixed source
row's outgoing weights are not constrained to sum to $1$.

\paragraph{$\varepsilon$-closure conditional normalisation.}
A more useful local property is the following.  Define
$w_X^L$ as the total WFST weight of paths starting at state $X$
that consume the next non-$\varepsilon$ input symbol $L$ (a character or
end-of-input sentinel $\$\,$), where each path is a (possibly empty)
geometric chain of insert self-loops followed by a single
input-consuming exit.  For our $6{\times}6$ WFST:
\begin{equation}
\begin{array}{c|cc}
X & w_X^c & w_X^\$ \\\hline
\sta   & 1 & 1 \\
\mat   & 1 & 1 \\
\ins_0 & 1 & 1 \\
\ins_1 & \kappa/p & 1/(1-\ext) \\
\del   & 1 & 1
\end{array}
\label{eq:wfst-eps-closure-table}
\end{equation}
The $\sta,\mat,\ins_0,\del$ rows are conditionally normalised: each
sums to exactly $1$.  The $\ins_1$ row alone is not, because its
destination divisors mix the $r$ (extension) and $1{-}r$ (exit) weights
asymmetrically into the single-step edges.

This local non-normalisation does \emph{not} affect the WFST's
correctness as a representation of $P(\text{descendant}\mid
\text{ancestor},\theta)$: every WFST path begins at $\sta$, and from
$\sta$ the chain-through-$\ins_1$ contribution accumulates exactly the
right $\kappa/p$ factor at the appropriate step to bring the total
back to $1$.  Concretely, starting from $\mat$ the $\varepsilon$-closure
weight to consume the next character is
\[
w_\mat^c
=\bigl(\tkfwfst''_{\mat\mat}+\tkfwfst''_{\mat\del}\bigr)
+\tkfwfst''_{\mat\ins_1}\cdot
   \frac{\tkfwfst''_{\ins_1\mat}+\tkfwfst''_{\ins_1\del}}{1-\tkfwfst''_{\ins_1\ins_1}}
=\frac{r+(1{-}r)(1{-}\beta)\kappa}{p}+(1{-}r)\beta\cdot\frac{\kappa}{p}
=\frac{r+(1{-}r)\kappa}{p}=1,
\]
so the $\kappa/p$ factor that the $\ins_1$ row alone undershoots is
exactly recovered when $\ins_1$ is reached from a ``proper'' source
state.  The $\ins_1$ row is improper only if the WFST were started
mid-stream from $\ins_1$ in isolation, which never happens in normal
use.

\paragraph{Per-input single-step row sums.}
The simpler invariant $\sum_b\tkfwfst''_{Xb}=1$ does \emph{not} hold
for any of the body rows.  This is a state-encoding artefact of folding
the Bernoulli-$r$ extension-vs-exit decision into the same outgoing
edges that carry the destination-singlet factor.  An alternative WFST
that exposes the extension event as a silent ``decision'' state would
be locally row-stochastic at the cost of more states.  The compact
$6{\times}6$ form trades local stochasticity for a smaller graph;
the global property
$\text{Singlet}\circ\tkfwfst''=\tkftrans''$
(with $\tkftrans''$ row-stochastic) is preserved either way.

\subsection{Testable Invariants}
\label{sec:wfst-invariants}

The following invariants hold for any valid parameters
$0 < \insrate, \delrate$, $\evoltime > 0$, $0 \leq \ext < 1$
and can be verified numerically from the WFST tensor code.

\begin{enumerate}[nosep]
\item \textbf{Fragment-boundary posterior:}
  $\ext/p + (1{-}\ext)\kappa/p = 1$.

\item \textbf{Pair HMM reconstruction:}
  For each state $a$ in $\{\sta,\mat,\ins_0,\ins_1,\del\}$, reconstruct
  the Pair HMM row by multiplying WFST entries by their singlet factors
  (per the rules above), and verify $\sum_b \tkftrans''_{ab} = 1$.

\item \textbf{Mixture decomposition:}
  $\tkfwfst''_{\mat\mat} = \ext/p + [(1{-}\ext)\kappa/p](1{-}\beta)\alpha$
  (and analogously for $\del \to \del$ with $\gamma$).
  Both mixture components are non-negative.

\item \textbf{TKF91 limit:}
  At $\ext = 0$: $p = \kappa$, the $\ins_0$ and $\ins_1$ rows
  become identical (and equal the TKF91 $\ins$ row), and the
  $6{\times}6$ WFST reduces to the TKF91 $5{\times}5$ WFST.

\item \textbf{$\fin$-column dichotomy:}
  $\tkfwfst''_{X\fin} = 1{-}\beta$ for $X \in \{\sta,\mat,\ins_1\}$
  and $1{-}\gamma$ for $X = \del$, but
  $\tkfwfst''_{\ins_0\fin} = (1{-}\ext)(1{-}\beta)$ scales linearly
  in $1{-}\ext$ because the $\ins_0$ exit weight is divided by the
  smaller $\sta\to\fin$ singlet weight $1{-}\kappa$ rather than
  $(1{-}\ext)(1{-}\kappa)$.

\item \textbf{$\ins_1$ column scaling:}
  $\tkfwfst''_{\mat\ins_1} = (1{-}\ext)\beta$ and
  $\tkfwfst''_{\del\ins_1} = (1{-}\ext)\gamma$ scale linearly in
  $(1{-}\ext)$, reflecting that inserts can only begin at fragment
  boundaries.

\item \textbf{Structural zeros:}
  $\tkfwfst''_{\sta\,\ins_1} = \tkfwfst''_{\mat\,\ins_0}
  = \tkfwfst''_{\ins_1\,\ins_0} = \tkfwfst''_{\del\,\ins_0} = 0$,
  reflecting that $\ins_0$ is reachable only before any ancestor
  has been consumed and $\ins_1$ only after.
\end{enumerate}

\subsection{Boundary-sampler factorisation of $\tkfwfst''$}
\label{sec:wfst92-boundary-factor}

The $6{\times}6$ WFST admits a clean factorisation into a boundary
sampler, a TKF91 WFST on fragment-level links, and a fragment-expansion
projection.  This makes the ``TKF91 hiding inside TKF92'' structure
explicit and exposes a degenerate Doob $h$-transform that is worth
recording.

\paragraph{Boundary sampler $B$.}  A row-stochastic WFST that reads
ancestor residues (without fragment markers) and outputs a per-residue
mark in $\{N, E\}$ ($N$ = ``new fragment,'' $E$ = ``extends current
fragment''):
\begin{itemize}[nosep]
\item From start $\sta_B$ on the first residue: emit $N$ with weight $1$.
\item From the working state $\mat_B$ on subsequent residues: emit $E$ with
  weight $r/p$, or $N$ with weight $(1-r)\kappa/p$.  Row sum on a
  residue input: $r/p + (1-r)\kappa/p = p/p = 1$.
\item EOS transitions to $\fin_B$ with weight $1$ from either state.
\end{itemize}

\paragraph{Trivial Doob $h$-transform.}  In TKF92 the residue identities
are i.i.d.\ $\sim \eqm$ {\em independent of\/} the boundary indicators,
so the posterior $P(\mathbf{b}|\mathbf{x})$ equals the prior
$P(\mathbf{b})$ residue-by-residue.  The Doob $h$-transform $h(b,x_i)
= P(x_i|b)/P(x_i)$ that would in general be required to make a
posterior boundary-sampler locally row-stochastic reduces to
$h \equiv 1$, leaving $B$ a strict prior sampler.  For models where
the per-residue emission distribution does depend on fragment-context
(some MixDom variants), $B$ would carry a non-trivial $h$ and would
no longer be locally row-stochastic without explicit normalisation.

\paragraph{Factorisation.}  Let $W_{91}^{\text{link}}$ denote the
TKF91 conditional WFST (Section~\ref{sec:tkf91-machines}) acting on
fragment-level links (one symbol per ancestor fragment-start), and
let $F$ denote the deterministic expand-and-erase projection that
inflates each TKF91 link into a $1{+}\text{Geom}(r)$-length
descendant fragment of residues (drawn from $\exp(Q\evoltime)$ for
match links, from $\eqm$ for insert links, or empty for delete links)
and discards the boundary marks from the output.  Then
\[
\tkfwfst'' \;=\; B \circ W_{91}^{\text{link}} \circ F .
\]

\paragraph{Row-by-row verification.}  The $\mat$-row entries of
$\tkfwfst''$ decompose as
\begin{eqnarray*}
\tkfwfst''_{\mat\,\mat}   & = & \underbrace{r/p}_{B:\,E} \cdot 1 + \underbrace{(1{-}r)\kappa/p}_{B:\,N} \cdot (1{-}\beta)\alpha \\
                          & = & \bigl(r + (1{-}r)(1{-}\beta)\kappa\alpha\bigr)/p \quad\checkmark \\
\tkfwfst''_{\mat\,\ins_1} & = & (1{-}r)\,\beta \quad \text{(new descendant link inserted between ancestor fragments)} \quad\checkmark \\
\tkfwfst''_{\mat\,\del}   & = & (1{-}r)\kappa/p \cdot (1{-}\beta)(1{-}\alpha) \;=\; (1{-}r)(1{-}\beta)\kappa(1{-}\alpha)/p \quad\checkmark \\
\tkfwfst''_{\mat\,\fin}   & = & 1{-}\beta \quad\checkmark
\end{eqnarray*}
The $\del$-row verifies identically with $\gamma$ in place of $\beta$.  The
$\ins_1$-row is the same TKF91 link-level row reached after some non-empty
ancestor prefix, weighted by the same $B$ structure as $\mat$.  The
$\sta$ and $\ins_0$ rows are the boundary case $b_1 = N$ (the first
residue is forced new), which removes the $r/p$ self-extend branch
and recovers the no-$r$ entries given in the table.

\paragraph{Detailed balance.}  Detailed balance holds at the augmented
$(x,\mathbf{b},y,\mathbf{b}')$ level: $W_{91}^{\text{link}}$ is reversible
under $\eqm$ (Holmes-Bruno), $B$ is a proper conditional sampler with
$h \equiv 1$, and marginalising the boundary marks $\mathbf{b}, \mathbf{b}'$
recovers the TKF92 joint pair HMM
$\tkftrans''$ exactly.

\paragraph{Consequence.}  For MCMC inference under TKF92 one can
augment the state with per-residue fragment-boundary indicators
$\mathbf{b}$ at each tree node (sampled cheaply from
$\text{Bernoulli}(r/p)$, {\em independently of the substitution
matrix\/}) and run TKF91 moves at the link level.  This decomposition
is the operational realisation of the ``fragments-as-TKF91-links''
view originally outlined in Thorne, Kishino, and Felsenstein (1992).
At $r = 0$ the $B$ step is trivial and the construction reduces to
plain TKF91.
